---
title: "Poland through the eyes of Eurostat Data"
author: "MK"
date: "April 2024"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
library(tidyverse)
library(extrafont)
library(svglite)
library(ggimage)
library(extrafont)
#font_import()
#loadfonts(device = "win")
#windowsFonts()

knitr::opts_chunk$set(
  echo = F,
  warning = F,
  collapse = T,
  message = F,
  dev = "svglite",
  fig.ext = ".svg"
)
options(scipen = 999)
```


# Energy

## Data Understanding

```{r}
searchData <- 
  eurostat::search_eurostat("Population")

data <- 
  eurostat::get_eurostat("env_ac_pefasu",
    type = "label", time_format = "num"
  )
```

```{r, out.width="100%", dpi = 300}
s <- svgstring()

data %>%
  mutate(NAs = if_else(is.na(values), "NA", "Value")) %>%
  group_by(geo, NAs) %>%
  summarise(n = n()) %>%
  group_by(geo) %>%
  mutate(height = if_else(NAs == "NA", sum(n), n)) %>%
  ggplot(
    aes(
      fill = NAs,
      y = reorder(geo, -n),
      x = n/1000
    )
  ) + 
  geom_col() + theme_minimal() +
  geom_text(aes(
    label = paste0(round(n/1000, 0), "k"), 
    y = reorder(geo, -n),
    x = height/1000 + 10
  ), inherit.aes = F, color = "#111d2b") +
  scale_fill_manual(values = c("#9fb1bc", "#2E4057")) +
  labs(
    x = " of records (thousends)",
    y = "",
    fill = "Missing data",
    title = "How many records are avaiable?",
    subtitle = "divided by a country / a group of countries",
    caption = "data source: eurostat \n dataset id:env_ac_pefasu"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 12),
    text = element_text(color = "#111d2b", family = "Segoe UI")
  )

htmltools::HTML(s())
invisible(dev.off())
```

```{r}
pop <- 
  eurostat::get_eurostat("demo_r_d2jan"
  )
pop %>%
  filter(
    sex == "T",
    age == "TOTAL"
  ) %>%
  filter(nchar(geo) == 2) %>%
  left_join(countrycode::codelist, join_by(geo == eurostat)) %>%
  mutate(
    id = paste0(iso.name.en, year(TIME_PERIOD))
  ) %>%
  select(id, "pop" = values) -> popeu
```

## Production 

```{r, out.width="100%", dpi = 300}
title <- stringr::str_wrap(
  "How does the polish energetic industry perform compare to other european countries with similar characteristics?"
)
subtitle <- stringr::str_wrap(
  "the latest period marked with a point"
)
s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "prod" = `Energy products`,
    geo
  ) %>% 
  filter(
    year %in% 2014:2022,
    geo != "European Union - 27 countries (from 2020)"
  ) %>%
  mutate(
    isPoland = if_else(geo == "Poland", "Poland", ""),
    id = paste0(geo, year)
  ) %>%
  left_join(popeu, join_by(id == id)) %>%
  mutate(
    pop = pop/1000000,
    prod = prod/1000000
  ) %>%
  filter(
    prod >= 2,
    pop >= 10
  ) %>%
  mutate(
    y2 = if_else(year == max(year, na.rm =T), prod, NA),
    x2 = if_else(year == max(year, na.rm =T), pop, NA)
  ) %>%
  ggplot() +
  geom_line(
    aes(
      y = prod,
      x = pop,
      color = isPoland,
      group = geo
    ), size = 1.1, show.legend = F
  ) +
  theme_minimal() +
    geom_point(
    aes(
      x = x2,
      y = y2
    ), color = "#111d2b", size = 2.2
  ) +
  geom_point(
    aes(
      x = if_else(isPoland == "Poland", x2, NA),
      y = if_else(isPoland == "Poland", y2, NA)
    ), color = "#d1000a", size = 2.2
  ) +
  geom_text(
    aes(
      x = x2,
      y = y2,
      label = geo
    ), nudge_x = 4
  ) +
  geom_abline(slope = 1/6, intercept = 0, linetype = "dashed", size = .5, color = "#3c3c3c") +
  scale_x_continuous(
    minor_breaks = seq(0, 90, 5),
    breaks = seq(0, 90, 15)
  ) +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  labs(
    x = "population (in milions of people)",
    y = "energy production (in milions of terajoules)",
    title = title,
    subtitle = "the latest period marked with a point",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) + 
   theme(
     panel.grid.minor = element_line(size = 0.25),
     panel.grid.major = element_line(size = 0.75),
     plot.title = element_text(face = "bold"),
     text = element_text(color = "#111d2b", family = "Segoe UI"),
     panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
     plot.background = element_rect(fill = '#fafafa', color = '#fafafa')
   ) 


htmltools::HTML(s())
invisible(dev.off())
```

1.  Compared to other European economies most similar to Poland in terms of population, Poles consume much less electricity per person than other countries - the only exception here is Italy. In turn, the economy with the highest electricity consumption per person is Germany.

2.  The more populated a country is, the more pronounced are the fluctuations caused by the COVID-19 outbreak. This may indicate the high industrialization of these countries - downtime in the production of goods resulted in a significant decline in electricity consumption. This thesis is confirmed in the case of the largest economies - German and French.

3.  Interestingly, none of the economies returned to their previous state of energy consumption in 2021, when restrictions and restrictions related to the pandemic were no longer so stringent.

## Waste

```{r, out.width="100%", dpi = 300}
s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "non-ren" = `Non-renewable waste`,
    "ren" = `Renewable waste`,
    "prod" = `Energy products`,
    geo
  ) %>% 
  mutate(
    isPoland = if_else(geo == "Poland", "Poland", ""),
    text = if_else(geo == "Poland" & year == 2020, "Poland", "")
  ) %>% na.omit %>%
  filter(
    ren != 0, `non-ren` != 0,
    geo != "European Union - 27 countries (from 2020)"
  ) %>%
  group_by(geo) %>%
  mutate(
    y2 = if_else(year == max(year, na.rm =T), `non-ren`, NA),
    x2 = if_else(year == max(year, na.rm =T), `ren`, NA)
  ) %>%
  ggplot(
    aes(
      x = ren,
      y = `non-ren`
    ), show.legend = F
  ) +
  geom_abline(xintercept = 0, size = .8, color = "#3c3c3c", linetype = "dashed", alpha = .9) +
  geom_line(
    aes(
      group = geo,
      color = isPoland,
      size = prod
    ), show.legend = F
  ) +
  geom_point(
    aes(
      x = x2,
      y = y2
    ), color = "#111d2b", size = 2.2
  ) +
  geom_point(
    aes(
      x = if_else(isPoland == "Poland", x2, NA),
      y = if_else(isPoland == "Poland", y2, NA)
    ), color = "#d1000a", size = 2.2
  ) +
  scale_size(range = c(.5, 2)) +
   geom_text(
     aes(
       x = x2,
       y = y2,
       label = if_else(geo != "Poland", geo, "")
     ), color = "#3c3c3c", alpha = 1, size = 3, nudge_y = .05
   ) +
  geom_text(
    aes(
      label = if_else(year == 2019, isPoland, "")
    ), color = "#F93943", alpha = .3, size = 5, nudge_x = -.2, nudge_y = .2
  ) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  labs(
    x = "Waste of renewable energy (Terajules)",
    y = "Waste of non-renewable energy (Terajules)",
    size = "Energy production",
    title = "Is Poland energy efficient?",
    subtitle = "compared to other eu countries in years 2008-2022;
    size idicates a total energy production, the latest period marked with a point",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) +
  theme_minimal() + 
  guides(
    color = "none",
    alpha = "none"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 12),
    text = element_text(color = "#111d2b", family = "Segoe UI")
  )


htmltools::HTML(s())
invisible(dev.off())
```



```{r, out.width="100%", dpi = 300}
title <- stringr::str_wrap(
  "How does Poland compare to other european countries with similar characteristics?"
)
subtitle <- stringr::str_wrap(
  "the latest period marked with a point"
)

s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "prod" = `Energy products`,
    geo
  ) %>% 
  filter(
    year %in% 2014:2022,
    geo != "European Union - 27 countries (from 2020)"
  ) %>% mutate(id = paste0(geo, year)) %>%
  left_join(popeu, join_by(id == id)) %>%
  group_by(geo) %>%
  summarise(
    prod = mean(prod, na.rm = T),
    pop = mean(pop, na.rm = T)
  ) %>%
  mutate(
    ratio = prod / pop,
    isPoland = if_else(geo == "Poland", "Poland", "No"),
    label = if_else(
      geo == "Poland",
      paste0(round(ratio, 2), " TJ per person"),
      round(ratio, 2) %>% as.character()
    )
  ) %>%
  na.omit %>%
  ggplot(aes(
      y = reorder(geo, ratio),
      x = ratio,
      fill = isPoland
    )
  ) + 
  geom_col(show.legend = F) + 
  geom_text(aes(label = label, color = isPoland), nudge_x = .05, hjust = 0, show.legend = F) +
  scale_fill_manual(values = c("#2E4057", "#F93943")) +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  theme_minimal() +
  labs(
    x = "Consumption in TJ / 1 person",
    y = "energy production (in milions of terajoules)",
    title = title,
    subtitle = subtitle,
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 12),
    text = element_text(color = "#111d2b", family = "Segoe UI")
  )

htmltools::HTML(s())
invisible(dev.off())
```
```{r, out.width="100%", dpi = 300}
s <- svgstring()

title <- stringr::str_wrap(
  "The average Pole consumes so much electricity per year that converting it into its caloric equivalent will result in 7 tons of chocolate."
)

subtitle <- stringr::str_wrap(
  "That is, 52 people (with an average requirement of 2000 kcal) could eat chocolate all year round!"
)

data.frame(
  x = rep(1:10, 7),
  y = rep(7:1, each = 10),
  img = c(rep(NA, 10), rep("www/icon.PNG", 52), rep(NA, 8))
) %>%
ggplot() +
  geom_image(
    aes(
      x = x,
      y = y,
      image = img
    ),
    size = .15
  ) +
  theme_void() + 
  labs(
    title = title,
    subtitle = subtitle
  ) +  
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 12),
    text = element_text(color = "#111d2b", family = "Segoe UI")
  ) 

htmltools::HTML(s())
invisible(dev.off())
```

## Sources
