---
title: "Poland through the eyes of Eurostat Data"
author: "MK"
date: "April 2024"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
library(tidyverse)
library(svglite)
library(ggimage)
library(extrafont)
library(plotly)
font_import()
loadfonts(device = "win")
windowsFonts()

knitr::opts_chunk$set(
  echo = F,
  warning = F,
  collapse = T,
  comment = F,
  out.width = "100%",
  dpi = 300,
  message = F,
  dev = "svglite",
  fig.ext = ".svg"
)
options(scipen = 999)
```

```{r}
searchData <- 
  eurostat::search_eurostat("Population")

data <- 
  eurostat::get_eurostat("env_ac_pefasu",
    type = "label", time_format = "num"
  )
```

```{r}
pop <- 
  eurostat::get_eurostat("demo_r_d2jan"
  )
pop %>%
  filter(
    sex == "T",
    age == "TOTAL"
  ) %>%
  filter(nchar(geo) == 2) %>%
  left_join(countrycode::codelist, join_by(geo == eurostat)) %>%
  mutate(
    id = paste0(iso.name.en, year(TIME_PERIOD))
  ) %>%
  select(id, "pop" = values) -> popeu
```

# Energy

The source of the analyzed data is the eurostat API from the `eurostat` package. As of the date of the analyses, the latest data cover the year 2022 for all EU countries.

## Data Understanding

The analyzed data set contains information on consumption, production and energy sources. Differentiating the industry (nace_r2 classification), country (limited to EU countries) and reporting period (2008-2022) with clear gaps in the continuity of reporting by individual countries. All values are in Terajoule units.

```{r, out.width="100%", dpi = 300}
s <- svgstring()

data %>%
  mutate(NAs = if_else(is.na(values), "NA", "Value")) %>%
  group_by(geo, NAs) %>%
  summarise(n = n()) %>%
  group_by(geo) %>%
  mutate(height = if_else(NAs == "NA", sum(n), n)) %>%
  ggplot(
    aes(
      fill = NAs,
      y = reorder(geo, -n),
      x = n/1000
    )
  ) + 
  geom_col() + theme_minimal() +
  geom_text(aes(
    label = paste0(round(n/1000, 0), "k"), 
    y = reorder(geo, -n),
    x = height/1000 + 10
  ), inherit.aes = F, color = "#111d2b") +
  scale_fill_manual(values = c("#9fb1bc", "#2E4057")) +
  labs(
    x = " of records (thousends)",
    y = "",
    fill = "Missing data",
    title = "How many records are avaiable?",
    subtitle = "divided by a country / a group of countries",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13)
  )

htmltools::HTML(s())
invisible(dev.off())

```

## Production

Since the subject of this analysis is Poland, I will start by comparing it in the context of other EU countries in terms of energy production and population. This will help position Poland in relation to other member states.

```{r, out.width="100%", dpi = 300}
title <- stringr::str_wrap(
  "How does the polish energetic industry perform compare to other european countries with similar characteristics?"
)
subtitle <- stringr::str_wrap(
  "the latest period marked with a point"
)
s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "prod" = `Energy products`,
    geo
  ) %>% 
  filter(
    year %in% 2014:2022,
    geo != "European Union - 27 countries (from 2020)"
  ) %>%
  mutate(
    isPoland = if_else(geo == "Poland", "Poland", ""),
    id = paste0(geo, year)
  ) %>%
  left_join(popeu, join_by(id == id)) %>%
  mutate(
    pop = pop/1000000,
    prod = prod/1000000
  ) %>%
  filter(
    prod >= 2,
    pop >= 10
  ) %>%
  mutate(
    y2 = if_else(year == max(year, na.rm = T), prod, NA),
    x2 = if_else(year == max(year, na.rm = T), pop, NA)
  ) %>%
  ggplot() +
  geom_line(
    aes(
      y = prod,
      x = pop,
      color = isPoland,
      group = geo
    ), size = 1.1, show.legend = F
  ) +
  theme_minimal() +
    geom_point(
    aes(
      x = x2,
      y = y2
    ), color = "#111d2b", size = 2.2
  ) +
  geom_point(
    aes(
      x = if_else(isPoland == "Poland", x2, NA),
      y = if_else(isPoland == "Poland", y2, NA)
    ), color = "#d1000a", size = 2.2
  ) +
  geom_text(
    aes(
      x = x2,
      y = y2,
      label = geo
    ), nudge_x = 4
  ) +
  geom_abline(slope = 1/6, intercept = 0, linetype = "dashed", size = .5, color = "#3c3c3c") +
  scale_x_continuous(
    minor_breaks = seq(0, 90, 5),
    breaks = seq(0, 90, 15)
  ) +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  labs(
    x = "population (in milions of people)",
    y = "energy production (in milions of terajoules)",
    title = title,
    subtitle = "the latest period marked with a point",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu; demo_r_d2jan"
  ) + 
   theme(
     panel.grid.minor = element_line(size = 0.25),
     panel.grid.major = element_line(size = 0.75),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13),
     panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
     plot.background = element_rect(fill = '#fafafa', color = '#fafafa')
   ) 


htmltools::HTML(s())
invisible(dev.off())
```

1.  Compared to other European economies most similar to Poland in terms of population, Poles consume much less electricity per person than other countries - the only exception here is Italy. In turn, the economy with the highest electricity consumption per person is Germany.

2.  The more populated a country is, the more pronounced are the fluctuations caused by the COVID-19 outbreak. This may indicate the high industrialization of these countries - downtime in the production of goods resulted in a significant decline in electricity consumption. This thesis is confirmed in the case of the largest economies - German and French.

3.  Interestingly, none of the economies returned to their previous state of energy consumption in 2022, when restrictions and restrictions related to the pandemic were no longer so stringent.

4.  Poland is one of the biggest energy producers, mainly due to the its population

However, these are very general and abstract sentences. Not very appealing to our heuristics and intuition. So let's look from the perspective of an average citizen.

```{r, out.width="100%", dpi = 300}
title <- stringr::str_wrap(
  "How does Poland compare to other european countries with similar characteristics?"
)
subtitle <- stringr::str_wrap(
  "the latest period marked with a point"
)

s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "prod" = `Energy products`,
    geo
  ) %>% 
  filter(
    year %in% 2014:2022,
    geo != "European Union - 27 countries (from 2020)"
  ) %>% mutate(id = paste0(geo, year)) %>%
  left_join(popeu, join_by(id == id)) %>%
  group_by(geo) %>%
  summarise(
    prod = mean(prod, na.rm = T),
    pop = mean(pop, na.rm = T)
  ) %>%
  mutate(
    ratio = prod / pop,
    isPoland = if_else(geo == "Poland", "Poland", "No"),
    label = if_else(
      geo == "Poland",
      paste0(round(ratio, 2), " TJ per person"),
      round(ratio, 2) %>% as.character()
    )
  ) %>%
  na.omit %>%
  ggplot(aes(
      y = reorder(geo, ratio),
      x = ratio,
      fill = isPoland
    )
  ) + 
  geom_col(show.legend = F) + 
  geom_text(aes(label = label, color = isPoland), nudge_x = .05, hjust = 0, show.legend = F) +
  scale_fill_manual(values = c("#2E4057", "#F93943")) +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  theme_minimal() +
  labs(
    x = "Consumption in TJ / 1 person",
    y = "energy production (in milions of terajoules)",
    title = title,
    subtitle = subtitle,
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13)
  )

htmltools::HTML(s())
invisible(dev.off())
```

1.  The average Pole does not consume (relatively) large amounts of energy. In this respect, it is definitely on the side of the eastern part of Europe, which closes this ranking.

2.  Of course, consumption also results directly from the industrialization of a given country - production markets will be largely responsible for the increase in this statistic.

If we want to convert 0.15TJ, which is still abstract for the average consumer, let's play a little with mathematics.

```{r, out.width="100%", dpi = 300}
s <- svgstring()

title <- stringr::str_wrap(
  "The average Pole consumes so much electricity per year that converting it into its caloric equivalent will result in 7 tons of chocolate."
)

subtitle <- stringr::str_wrap(
  "That is, 52 people (with an average requirement of 2000 kcal) could eat chocolate all year round!"
)

data.frame(
  x = rep(1:10, 7),
  y = rep(7:1, each = 10),
  img = c(rep(NA, 10), rep("www/icon.PNG", 52), rep(NA, 8))
) %>%
ggplot() +
  geom_image(
    aes(
      x = x,
      y = y,
      image = img
    ),
    size = .15
  ) +
  theme_void() + 
  labs(
    title = title,
    subtitle = subtitle, 
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) +  
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13)
  ) 

htmltools::HTML(s())
invisible(dev.off())
```

After converting energy consumption into kcal, assuming the caloric value of chocolate at the level of 546kcal/100g, it turns out that the average Pole consumes the equivalent of electricity, which would allow for 7 tons of chocolate - that is, feeding 52 people for a whole year! Moving on to more realistic calculations, the same amount of energy would be sufficient for [Tesla Model X](https://ev-lectron.com/blogs/blog/how-much-electricity-does-it-take-to-charge-a-tesla#:~:text=The%20capacity%20of%20Tesla%20electric,Model%20S%20and%20X%20variants) circle the Earth along the equator approximately **6,372 times**! Or to go on **the moon 664 times** :)

Now that we know how much energy we use, it's time to think about what we actually use?

```{r, out.width="100%"}
s <- svgstring()
pal <- RColorBrewer::brewer.pal(name = "Oranges", n = 9)[4:9]

data %>% filter(
  prod_nrg %in% c(
    "Hard coal",
    "Brown coal and peat",
    "Derived gases (= manufactured gases excl. biogas)",
    "Secondary coal products (coke, coal tar, patent fuel, BKB and peat products)",
    "Crude oil, NGL, and other hydrocarbons (excl. bio)",
    "Natural gas (without bio)",
    "Motor spirit (without bio)",
    "Kerosenes and jet fuels (without bio)",
    "Naphtha",
    "Transport diesel (without bio)",
    "Heating and other gasoil (without bio)",
    "Residual fuel oil",
    "Refinery gas, ethane and LPG",
    "Other petroleum products incl. additives/oxygenates and refinery feedstocks",
    "Nuclear fuel",
    "Wood, wood waste and other solid biomass, charcoal"   
  ),
  geo == "Poland",
  stk_flow == "Use",
  nace_r2 == "Total - all NACE activities"
) %>%
  group_by(TIME_PERIOD) %>%
  mutate(
    all = sum(values),
    perc = values / sum(values),
    prod_nrg = if_else(perc >= .05, prod_nrg, "Others")
  ) %>%
  group_by(TIME_PERIOD, prod_nrg) %>% summarise(values = sum(values)) %>%
  ggplot(
    aes(
      x = TIME_PERIOD,
      y = values,
      color = prod_nrg,
      group = prod_nrg
    )
  ) +
  geom_line() +
  geom_text(aes(
    label = paste0(format(round(values), big.mark = "'", scientific = F), "Tj"),
    x = if_else(TIME_PERIOD == 2021, TIME_PERIOD, NA),
  ), hjust = 1, nudge_y = 50000) +
  geom_text(aes(
    label = prod_nrg,
    x = if_else(TIME_PERIOD == 2014, TIME_PERIOD, NA),
  ), hjust = 0, nudge_y = 50000) +
  theme_minimal() +
  scale_x_continuous(breaks = 2014:2021) + 
  guides(
    color = "none",
    alpha = "none"
  ) + 
  scale_color_manual(values = pal) +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13)
  ) +
  labs(
    x = "",
    y = "Energy source (Terajules)",
    title = "What drives Polish development?",
    subtitle = "energy source with stake below 5% classified as others",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) 

htmltools::HTML(s())
invisible(dev.off())
```

No major surprise - Poland consumes most of the energy from hard coal. An interesting case is also natural gas, which is characterized by high variability of use over the years. This may indicate problems with the use (power plants) or delivery of this raw material to Poland.

However, returning to coal - is it an exception on a global scale; is Poland extremely carbon-intensive?

```{r, out.width="100%", out.height=600}
# light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = F,
  showcoastlines = F,
  projection = list(type = 'Mercator'),
  plot_bgcolor = "rgb(250, 250, 250)",
  paper_bgcolor = "rgb(250, 250, 250)",
  fig_bgcolor = "rgb(250, 250, 250)",
  scope = 'europe',
 lonaxis = list(range = c(-30, 60)),
 lataxis = list(range = c(40, 70))
)

data %>%
  filter(
    prod_nrg %in% c(
      "Hard coal",
      "Brown coal and peat",
      "Derived gases (= manufactured gases excl. biogas)",
      "Secondary coal products (coke, coal tar, patent fuel, BKB and peat products)",
      "Crude oil, NGL, and other hydrocarbons (excl. bio)",
      "Natural gas (without bio)",
      "Motor spirit (without bio)",
      "Kerosenes and jet fuels (without bio)",
      "Naphtha",
      "Transport diesel (without bio)",
      "Heating and other gasoil (without bio)",
      "Residual fuel oil",
      "Refinery gas, ethane and LPG",
      "Other petroleum products incl. additives/oxygenates and refinery feedstocks",
      "Nuclear fuel",
      "Wood, wood waste and other solid biomass, charcoal"   
    ),
    TIME_PERIOD == 2021,
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities"
  ) %>%
  group_by(prod_nrg) %>%
  mutate(
    values = round(values / sum(values) * 100)
  ) %>%
  filter(prod_nrg == "Hard coal") %>%
  left_join(
    countrycode::codelist %>% select(iso.name.en, iso3c),
    join_by(geo == `iso.name.en`)
  ) %>%
  plot_geo() %>% 
  add_trace(
    z = ~values,
    color = ~values,
    colors = 'Blues',
    text = ~geo,
    locations = ~iso3c,
    marker = list(line = l)
  ) %>%
  colorbar(title = 'Hard Coal use', ticksuffix = '%', limits = c(0, 16)) %>%
  layout(
    title = "",
    caption = "asd",
    geo = g
  )  %>%
  layout(
    margin = list(l = 50, r = 50, b = 100, t = 50),
    annotations = list(
      x = 1, y = -0.3, text = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu",
      xref = 'paper', yref = 'paper', showarrow = F,
      xanchor = 'right', yanchor = 'auto', xshift = 50, yshift = 20,
      font = list(size = 10)
    ),
      plot_bgcolor = "rgb(250, 250, 250)",
      paper_bgcolor = "rgb(250, 250, 250)",
      fig_bgcolor = "rgb(250, 250, 250)"
  )
```

As the map above shows, Poland is the leader in this infamous ranking. Interestingly, it is followed by Germany - with 12% of energy sources in the form of hard coal.

## Waste

Knowing where Poland stands on the European stage in terms of production, I will now look at how much this production is used - this will be an introduction to the study of how Poland (but also other countries) manage their energy.

```{r, out.width="100%", dpi = 300}
s <- svgstring()

data %>%
  filter(
    stk_flow == "Use",
    nace_r2 == "Total - all NACE activities",
    prod_nrg %in% c(
      "Non-renewable waste",
      "Renewable waste",
      "Energy products"
    )
  ) %>% 
  pivot_wider(values_from = "values", names_from = "prod_nrg") %>%
  select(
    "year" = TIME_PERIOD,
    "non-ren" = `Non-renewable waste`,
    "ren" = `Renewable waste`,
    "prod" = `Energy products`,
    geo
  ) %>% 
  mutate(
    isPoland = if_else(geo == "Poland", "Poland", ""),
    text = if_else(geo == "Poland" & year == 2020, "Poland", "")
  ) %>% na.omit %>%
  filter(
    ren != 0, `non-ren` != 0,
    geo != "European Union - 27 countries (from 2020)"
  ) %>%
  group_by(geo) %>%
  mutate(
    y2 = if_else(year == max(year, na.rm = T), `non-ren`, NA),
    x2 = if_else(year == max(year, na.rm = T), `ren`, NA)
  ) %>%
  ggplot(
    aes(
      x = ren,
      y = `non-ren`
    ), show.legend = F
  ) +
  geom_abline(xintercept = 0, size = .8, color = "#3c3c3c", linetype = "dashed", alpha = .9) +
  geom_line(
    aes(
      group = geo,
      color = isPoland,
      size = prod
    ), show.legend = F
  ) +
  geom_point(
    aes(
      x = x2,
      y = y2
    ), color = "#111d2b", size = 2.2
  ) +
  geom_point(
    aes(
      x = if_else(isPoland == "Poland", x2, NA),
      y = if_else(isPoland == "Poland", y2, NA)
    ), color = "#d1000a", size = 2.2
  ) +
  scale_size(range = c(.5, 2)) +
   geom_text(
     aes(
       x = x2,
       y = y2,
       label = if_else(geo != "Poland", geo, "")
     ), color = "#3c3c3c", alpha = 1, size = 3, nudge_y = .05
   ) +
  geom_text(
    aes(
      label = if_else(year == 2019, isPoland, "")
    ), color = "#F93943", alpha = 1, size = 3, nudge_y = .05
  ) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("#2E4057", "#F93943")) +
  labs(
    x = "Waste of renewable energy (Terajules)",
    y = "Waste of non-renewable energy (Terajules)",
    size = "Energy production",
    title = "Is Poland wasting its energy?",
    subtitle = "compared to other eu countries in years 2008-2022;
    size idicates a total energy production, the latest period marked with a point",
    caption = "Author: MK \n data source: eurostat \n dataset id:env_ac_pefasu"
  ) +
  theme_minimal() + 
  guides(
    color = "none",
    alpha = "none"
  ) + 
  theme(
    panel.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.background = element_rect(fill = '#fafafa', color = '#fafafa'),
    plot.title = element_text(face = "bold", size = 18),
    text = element_text(color = "#111d2b", family = "Segoe UI", size = 13)
  )


htmltools::HTML(s())
invisible(dev.off())
```

1.  In the ranking of energy waste, the top has undergone some changes, while Poland still holds a high position, interestingly - in recent years there has been a tendency to waste energy from renewable sources - this may be a consequence of a wild energy revolution, the potential of which is wasted due to lack of energy storage and, in general, inadequate infrastructure.

2.  Even though Poland is ahead of countries such as Spain or Italy in this infamous ranking, it is not (at least based on these data) in the worst situation. Scandinavian countries and the Netherlands, known for their extensive use of renewable energy, are doing much worse.
